---
title: "CEL file playground"
author: "Erich Robbi"
date: "2022-11-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preparation of the required libraries

Install BiocManager

```{r}

if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

# BiocManager::install("affyio")

```

Install the required packages

```{r}
BiocManager::install("affy")
BiocManager::install("hgu133plus2cdf")
BiocManager::install("hgu133plus2.db")
BiocManager::install("limma")
BiocManager::install("annotate")
BiocManager::install("lumiHumanAll.db")
BiocManager::install("GEOquery")

```

Load the libraries

```{r}
library(affy)
library(limma)
library(annotate)
library(hgu133plus2.db)
library(lumiHumanAll.db)
library(biomaRt)
library(tidyverse)
library(GEOquery)
```

# Retrieve data

Read the input data

```{r}
files <- list.files("../data/", full.names = TRUE)
data3 <- ReadAffy(filenames = files)
```

Normalization and expression matrix

```{r}
rma.data <- affy::rma(data3)
expression.matrix <- exprs(rma.data)

# has.cancer <- rep(0, 121)
# has.cancer[18:121] <- 1
# exp2 <- as_tibble(expression.matrix)
# exp2[nrow(exp2) + 1,] = t(has.cancer)
# exp2$genes <- c(row.names(expression.matrix), 'has_cancer')
# write.csv(exp2, 'expr_clark.csv', row.names = F)
```

```{r}
rma.medians <- rowMedians(Biobase::exprs(rma.data))
hist_res <- hist(rma.medians, 100, col = "cornsilk1", freq = FALSE, 
            main = "Histogram of the median intensities", 
            border = "antiquewhite4",
            xlab = "Median intensities")
```


we read series matrix file
```{r}
gse=getGEO(filename="GSE45827_series_matrix.txt")
```

We extract information from it
```{r}
i <- 1
type_cancer <- c()
diagnosis <- c()
sample_accession <- c()
source_name <- c()
while(i <= length(gse$title)){
  if(str_detect(gse$title[i], "rep1-2")){
    type_cancer <- c(type_cancer, rep(gse$`tumor subtype:ch1`[i], 2))
    diagnosis <- c(diagnosis, rep(gse$characteristics_ch1[i], 2))
    sample_accession <- c(sample_accession, rep(gse$geo_accession[i], 2))
    source_name <- c(source_name, rep(gse$source_name_ch1[i], 2))
  }
  else{
    type_cancer <- c(type_cancer, gse$`tumor subtype:ch1`[i])
    diagnosis <- c(diagnosis, gse$characteristics_ch1[i])
    sample_accession <- c(sample_accession, gse$geo_accession[i])
    source_name <- c(source_name, gse$source_name_ch1[i])
  }
  i <- i + 1
}
```

```{r}
for(i in 1:length(diagnosis)){
  diagnosis[i] <- str_replace(diagnosis[i], "diagnosis: ", "")
}
```


We merge the information with the expression matrix and save the csv
```{r}
exp_df <- as_tibble(t(expression.matrix))
exp_df$type <- type_cancer
exp_df$diagnosis <- diagnosis
exp_df$sample <- sample_accession
exp_df$source <- source_name

write.csv(exp_df, "expr_new.csv")
```


Using the biomaRt package, we retrieve ensembl_gene_id, gene symbols, and annotation information from the Ensembl database. Notice that with attributes you specify the information you want to retrieve from Ensembl. Instead, with filter you specify what you have to initiate the research in the database.

```{r}

mart <- useMart(dataset = "hsapiens_gene_ensembl", biomart='ensembl')

mart <- useDataset("hsapiens_gene_ensembl", mart)


row_ids <- row.names(expression.matrix)

annotLookup <- getBM(
  mart = mart,
  attributes = c(
    "ensembl_gene_id",
    "affy_hg_u133_plus_2",
    "hgnc_symbol", "description","refseq_mrna", "refseq_ncrna"),
  filter = "affy_hg_u133_plus_2",
  values = row_ids,
  uniqueRows=TRUE)

write.csv(annotLookup, 'annotation.csv', row.names = F)
```

To get all the possible attributes we can specify, we can use the following command.

```{r}
listAttributes(mart)
```

# Check

Now, we check if the dataset contains the snoRNAs of interest.

```{r}
snorna_annotati <- annotLookup[str_detect(annotLookup$hgnc_symbol,"SNOR")== TRUE & annotLookup$affy_hg_u133_plus_2 != "",]
```

```{r}
gene.set.of.interest <- c("SNORA1","SNORA12","SNORA14B","SNORA16A","SNORA21","SNORA23","SNORA24","SNORA32","SNORA38","SNORA44","SNORA48","SNORA49","SNORA52","SNORA53","SNORA57","SNORA61","SNORA63","SNORA64","SNORA65","SNORA70","SNORA71C","SNORA73A","SNORA73B","SNORA75","SNORA78","SNORA8","SNORD10","SNORD102","SNORD104","SNORD105","SNORD105B","SNORD108","SNORD110","SNORD111B","SNORD113-3","SNORD113-4","SNORD114-1","SNORD114-13","SNORD114-14","SNORD114-19","SNORD114-20","SNORD114-21","SNORD115-23","SNORD115-32","SNORD116-13","SNORD119","SNORD12","SNORD12B","SNORD13","SNORD14A","SNORD14C","SNORD14D","SNORD15B","SNORD16","SNORD17","SNORD18A","SNORD1B","SNORD20","SNORD22","SNORD26","SNORD28","SNORD29","SNORD32A","SNORD33","SNORD34","SNORD35A","SNORD36A","SNORD36B","SNORD38A","SNORD3A","SNORD3D","SNORD41","SNORD42A","SNORD42B","SNORD44","SNORD45A","SNORD47","SNORD49B","SNORD4B","SNORD5","SNORD50A","SNORD52","SNORD53","SNORD54","SNORD55","SNORD56","SNORD57","SNORD58A","SNORD58C","SNORD59A","SNORD60","SNORD63","SNORD64","SNORD65","SNORD68","SNORD69","SNORD71","SNORD74","SNORD76","SNORD8","SNORD81","SNORD82","SNORD84","SNORD86","SNORD87","SNORD89","SNORD94","SNORD96A","SNORD97","SNORD99","TMX1","ZFAS1","CDKN2B-AS1","CWF19L1","EIF4A1","EP400","HIF1A-AS2","MEG8","NOP56","RACK1","RPL13A","RPS13","SCARNA12","SNHG12","SNHG20","SNHG5","SNHG6","SNHG7","SNHG8","AP1G1","CFDP1","CHD8","DDX39B","EEF2","EIF4A2","EIF4G2","GAS5","GNL3","HSPA8","HSPA9","IPO7","MYRIP","NAN","NCL","NFATC3","PCAT4","PPAN","PRKAA1","PRRC2A","PTCD3","RABGGTB","RNF149","RPL10","RPL12","RPL13","RPL17","RPL21","RPL23","RPL23A","RPL4","RPL7A","RPLP2","RPS11","RPS2","RPS20","RPS3","RPS8","SF3B3","SLC25A3","SNHG1","SNHG3","SNHG9","SNORD1C","SNORD35B","SNORD37","SNRPB","SNX5","TAF1D","TNPO2","TOMM20","WDR43")
```

```{r}
genes.in.dataset <- annotLookup[annotLookup$hgnc_symbol %in% gene.set.of.interest,]

genes.in.dataset.no.duplicates <- unique(genes.in.dataset$hgnc_symbol)
print("dimension of our gene set:")
length(gene.set.of.interest)
print("genes of interest present in the dataset:")
length(genes.in.dataset.no.duplicates)
```

We can now label the data according to the samples in the links at https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6090421/table/tI-etm-0-0-6268/?report=objectonly
```{r}
df <- t(expression.matrix)
df <- as_tibble(df)

# 0 = no cancer
# 1 = cancer

has.cancer <- rep("NoCancer", 121)
has.cancer[18:121] <- "Cancer"
df$cancer <- has.cancer
```

To initially reduce the number of variables used, we can select only the relevant genes for our project
```{r}
df_filtered <- subset(df, select=c(genes.in.dataset$affy_hg_u133_plus_2))
df_filtered$cancer <- has.cancer
```

The new dataset has 731 variables instead of more than 54000. Nice. Now we can try our hand at classification:
```{r}
library(caret)
set.seed(16112022)
sample <- caret::createDataPartition(df_filtered$cancer, p=0.8)
train_data <- df_filtered[sample$Resample1, ]
test_data <- df_filtered[-sample$Resample1, ]
ctrl <- trainControl(method="cv")
```

```{r}
fitNaive <- train(cancer ~ .,
                  data = train_data,
                  method = "naive_bayes",
                  trControl = ctrl)
predNaive <- predict(fitNaive, test_data, type = "prob")
predFinal <- ifelse(predNaive[,2] >= 0.5, "Cancer", "NoCancer")
table(predFinal, test_data$cancer)
```

There is an error! Upon further inspection, many columns have the same name because they are the same gene (but with different values). Let's change the column names:
```{r}
library(collections, warn.conflicts = FALSE)

columns_filtered <- colnames(df_filtered)
new_columns <- c()
previous_columns <- dict()
previous_count <- 1
for (column in columns_filtered) {
  if (previous_columns$has(column)){
    previous_columns$set(column, previous_columns$get(column) + 1)
  }
  else{
    previous_columns$set(column, 1)
  }
  if(column == "cancer"){
    new_columns <- c(new_columns, column)
  }
  else{
    new_columns <- c(new_columns, paste0(column, "_", previous_columns$get(column)))
  }
}

colnames(df_filtered) <- new_columns
```

Now we can retry our classification:
```{r}
set.seed(16112022)
sample <- caret::createDataPartition(df_filtered$cancer, p=0.8)
train_data <- df_filtered[sample$Resample1, ]
test_data <- df_filtered[-sample$Resample1, ]

fitNaive <- train(as.factor(cancer) ~ .,
                  data = train_data,
                  method = "naive_bayes",
                  trControl = ctrl)
predNaive <- predict(fitNaive, test_data, type = "prob")
predFinal <- ifelse(predNaive[,2] >= 0.5, "NoCancer", "Cancer")
table(predFinal, test_data$cancer)
```

```{r}
test_data$cancer
```

```{r}
ifelse(predNaive[,2] >= 0.5, "NoCancer", "Cancer")
```

```{r}
mean(predFinal == test_data$cancer)
```

```{r}
sensitivity(table(predFinal,test_data$cancer))
```

```{r}
specificity(table(predFinal,test_data$cancer))
```

